#Personal Activity Monitoring Devices Data Analysis

First, we will read the data from the Activity Monitoring Data CSV: 
```{r firstchunk, echo=TRUE}
data <- read.csv("activity.csv", na.strings="NA")
```

We will then remove the NAs from the dataset prior to the analysis: 

```{r secondchunk, echo=TRUE}
datanoNA <- data[!is.na(data$steps),]
```

##Number of Steps Taken Each Day

First, we need to sum the amount of steps taken each day and create a new data table with the results:

```{r thirdchunk, echo=TRUE}
aggdata <- aggregate(steps~date,datanoNA,sum)
```

A histogram is created showing the number of steps taken each day: 
```{r fourthchunk, echo = TRUE}
hist(aggdata$steps,xlab="Steps Taken Per Day",main="Total Number of Steps Taken Each Day")
```

###Mean Number of Steps 

The mean for the total number of steps per day is:
```{r fifthchunk, echo= TRUE}
mean1 <- mean(aggdata$steps)
mean1
```

###Median Number of Steps

The median for the total number of steps per day is:
```{r sixthchunk, echo= TRUE}
median1 <- median(aggdata$steps)
median1
```

##Average Number of Steps At Each 5-Minute Interval

Now, we will create a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days.  First, we will create a new data table with the average number of steps taken for each interval: 

```{r seventhchunk, echo=TRUE}
aggdata2 <- aggregate(steps~interval,datanoNA,mean)
```

The following is the plot for this step.  The x-axis labels mean the following: 500 = hour 5, minute 0; 1005 = hour 10, minute 5, etc: 

```{r eighthchunk, echo=TRUE}
plot.ts(aggdata2$interval,aggdata2$steps,type="l",xlab="5-Minute Interval",ylab="Average Number of Steps",main="Average Number of Steps for Each 5-Minute Interval")
```

###Interval Where Maximum Number of Steps Was Taken

The 5 minute variable that contains the maximum number of steps is:
```{r ninthchunk, echo=TRUE}
with(aggdata2, interval[steps==max(steps)])
```

This corresponds to 8:35am.  

##Missing Values Analysis

Now, we will work with missing values.  The total number of missing values in the original dataset is:

```{r tenthchunk, echo=TRUE}
na <- data[!complete.cases(data),]
nrow(na)
```

Since missing values may introduce bias into our analyses, we will now insert values in place of the NAs in the original dataset.  To do this, I will use the mean of each interval from the aggdata2 datatable:

```{r eleventhchunk, echo=TRUE}
data.na <- data$interval[is.na(data$steps)]
value <- sapply(data.na, function(steps) which (aggdata2$interval == steps))
data$steps[is.na(data$steps)] <- aggdata2$steps[value]
data
```

###Number of Steps Taken Each Day From New Full Dataset

Create a new histogram of the total number of steps taken each day:

First, we need to again sum the amount of steps taken each day and create a new data table with the results:

```{r twelfthchunk, echo=TRUE}
fullaggdata <- aggregate(steps~date,data,sum)
```

A histogram is created showing the number of steps taken each day with the new full dataset:

```{r thirteenthchunk, echo = TRUE}
hist(fullaggdata$steps,xlab="Steps Taken Per Day",main="Total Number of Steps Taken Each Day")
```

###Mean of Total Number of Steps

The mean for the total number of steps per day from the new full dataset is:
```{r fourteenthchunk, echo= TRUE}
mean2 <- mean(fullaggdata$steps)
mean2
```

###Median of Total Number of Steps

The median for the total number of steps per day from the new full dataset is:
```{r fifteenthchunk, echo= TRUE}
median2 <- median(fullaggdata$steps)
median2
```

##Analysis of Dataset Including Values in NAs versus the Dataset from Above Excluding NA Values

We will now analyze the differences in the means and medians from the old and new datasets to see if adding values to the NAs actually made a difference.  To do this, we will take the difference between the old and new means and medians:
```{r sixteenthchunk, echo=TRUE}
meandif <- mean2-mean1
#Mean difference
meandif

mediandif <- median2-median1
##Median difference
mediandif
```

From these differences, it actually looks like adding in averages for the NA values did not have much of an impact.

##Weekday vs. Weekend Activity

Our final analysis will involve looking at activity patterns on weekdays versus weekends from the dataset with the filled-in missing values.

First, we need to create a new variable that has two levels ("weekday" or "weekend"), corresponding to whether a given date is a weeday or weekend day.

```{r seventeenthchunk, echo=TRUE}
data$type <- weekdays(data$date, abbreviate=FALSE)
data$type[data$type=="Monday"] <- "weekday"
data$type[data$type=="Tuesday"] <- "weekday"
data$type[data$type=="Wednesday"] <- "weekday"
data$type[data$type=="Thursday"] <- "weekday"
data$type[data$type=="Friday"] <- "weekday"
data$type[data$type=="Saturday"] <- "weekend"
data$type[data$type=="Sunday"] <- "weekend"
```

Finally, we will create a panel plot with a time series plot of the 5-minute interval and average number of steps taken, averaged across all weekday days and weekend days.  

```{r eighteenthchunk, echo=TRUE}
aggdata3 <- aggregate(steps~interval + type,data,mean)
```

The following is the plot for this step.  The x-axis labels mean the following: 500 = hour 5, minute 0; 1005 = hour 10, minute 5, etc: 

```{r nineteenthchunk, echo=TRUE,fig.height=4}
##dataset with only weekday data
aggdata4 <- aggdata3[aggdata3$type=="weekday",]
##dataset with only weekend data
aggdata5 <- aggdata3[aggdata3$type=="weekend",]
##plot
par(mfrow=c(1,2),mar=c(5,4.5,4.5,4.5))
plot.ts(aggdata4$interval,aggdata4$steps,type="l",xlab="5 Minute Interval",ylab="Average Number of Steps",main="Average Number of Steps For Each 5-Minute Interval on Weekdays")
plot.ts(aggdata5$interval,aggdata5$steps,type="l",xlab="5 Minute Interval",ylab="Average Number of Steps",main="Average Number of Steps For Each 5-Minute Interval on Weekends")
```