#Personal Activity Monitoring Devices Data Analysis

In this report, we will be analyzing an individual's personal activity monitoring device.  The device totaled the number of steps taken by the individual at 5 minute intervals throughout the day.  Our dataset contains data from October 2012 through November 2012.  We will first calculate the mean and median number of the number of steps ignoring missing values in the dataset (which are designated as NA).  Next, we will take the average number of steps taken in each 5 minute interval across all days.  Next, we will input values for the missing values in the dataset, recalculate the mean and median, and compare this to the other calculations we made ignoring the NAs.  Finally, we will use the complete dataset to analyze activity on weekdays and weekends.

First, we will read the data from the Activity Monitoring Data CSV.  The file was downloaded on June 8th at 12:08pm EST.  
```{r firstchunk, echo=TRUE}
data <- read.csv("activity.csv", na.strings="NA")
## convert the date variable into a Date class instead of factor
data$date <- as.Date(data$date, format="%Y-%m-%d")
```

We will then remove the NAs from the dataset for the first part of our analysis: 

```{r secondchunk, echo=TRUE}
datanoNA <- data[!is.na(data$steps),]
```

##Number of Steps Taken Each Day

Now, we can begin analysis.  We will start by looking at the average number of steps taken each day.  To do this, first, we need to sum the amount of steps taken each day and create a new data table with the results:

```{r thirdchunk, echo=TRUE}
aggdata <- aggregate(steps~date,datanoNA,sum)
```

A histogram is created showing the number of steps taken each day: 
```{r fourthchunk, echo = TRUE}
hist(aggdata$steps,xlab="Steps Taken Per Day",main="Total Number of Steps Taken Each Day")
```

###Mean Number of Steps 

The mean for the total number of steps per day is:
```{r fifthchunk, echo= TRUE}
mean1 <- mean(aggdata$steps)
mean1
```

###Median Number of Steps

The median for the total number of steps per day is:
```{r sixthchunk, echo= TRUE}
median1 <- median(aggdata$steps)
median1
```

##Average Number of Steps At Each 5-Minute Interval

Now, we will create a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days.  First, we will create a new data table with the average number of steps taken for each interval: 

```{r seventhchunk, echo=TRUE}
aggdata2 <- aggregate(steps~interval,datanoNA,mean)
```

The following is the plot showing the average number of steps taken, averaged across all days.  The x-axis labels can be interpreted as the following: 500 = hour 5, minute 0; 1005 = hour 10, minute 5, etc: 

```{r eighthchunk, echo=TRUE}
plot.ts(aggdata2$interval,aggdata2$steps,type="l",xlab="5-Minute Interval",ylab="Average Number of Steps",main="Average Number of Steps for Each 5-Minute Interval")
```

###Interval Where Maximum Number of Steps Was Taken

The 5 minute variable that contains the maximum number of steps is:
```{r ninthchunk, echo=TRUE}
with(aggdata2, interval[steps==max(steps)])
```

This corresponds to 8:35am.  

##Missing Values Analysis

Now, we will work with missing values.  The total number of missing values in the original dataset is:

```{r tenthchunk, echo=TRUE}
na <- data[!complete.cases(data),]
nrow(na)
```

Since missing values may introduce bias into our analyses, we will now insert values in place of the NAs in the original dataset.  To do this, I will use the mean of each interval from the aggdata2 datatable:

```{r eleventhchunk, echo=TRUE}
data.na <- data$interval[is.na(data$steps)]
value <- sapply(data.na, function(steps) which (aggdata2$interval == steps))
data$steps[is.na(data$steps)] <- aggdata2$steps[value]
```

###Number of Steps Taken Each Day From New Full Dataset

We will now create a new histogram of the total number of steps taken each day.

Like with the first dataset, first, we need to again sum the amount of steps taken each day and create a new data table with the results:

```{r twelfthchunk, echo=TRUE}
fullaggdata <- aggregate(steps~date,data,sum)
```

A histogram is created showing the number of steps taken each day with the new full dataset:

```{r thirteenthchunk, echo = TRUE}
hist(fullaggdata$steps,xlab="Steps Taken Per Day",main="Total Number of Steps Taken Each Day")
```

###Mean of Total Number of Steps

The mean for the total number of steps per day from the new full dataset is:
```{r fourteenthchunk, echo= TRUE}
mean2 <- mean(fullaggdata$steps)
mean2
```

###Median of Total Number of Steps

The median for the total number of steps per day from the new full dataset is:
```{r fifteenthchunk, echo= TRUE}
median2 <- median(fullaggdata$steps)
median2
```

##Analysis of Dataset Including Values in NAs versus the Dataset from Above Excluding NA Values

We will now analyze the differences in the means and medians from the old and new datasets to see if adding values to the NAs actually made a difference.  To do this, we will take the difference between the old and new means and medians.

The first answer shows the difference in means; the second reveals the difference in medians:
```{r sixteenthchunk, echo=TRUE}
meandif <- mean2-mean1
#Mean difference
meandif

mediandif <- median2-median1
##Median difference
mediandif
```

From these differences, it actually looks like adding in averages for the NA values did not have much of an impact.

##Weekday vs. Weekend Activity

Our final analysis will involve looking at activity patterns on weekdays versus weekends from the dataset with the filled-in missing values.

First, we need to create a new variable that has two levels ("weekday" or "weekend"), corresponding to whether a given date is a weeday or weekend day.

```{r seventeenthchunk, echo=TRUE}
data$type <- weekdays(data$date, abbreviate=FALSE)
data$type[data$type=="Monday"] <- "weekday"
data$type[data$type=="Tuesday"] <- "weekday"
data$type[data$type=="Wednesday"] <- "weekday"
data$type[data$type=="Thursday"] <- "weekday"
data$type[data$type=="Friday"] <- "weekday"
data$type[data$type=="Saturday"] <- "weekend"
data$type[data$type=="Sunday"] <- "weekend"
```

Finally, we will create a panel plot with a time series plot of the 5-minute interval and average number of steps taken, averaged across all weekday days and weekend days.  The dataset needed to create the plot is as follows.  It averages the number of steps taken along each interval.   

```{r eighteenthchunk, echo=TRUE}
aggdata3 <- aggregate(steps~interval + type,data,mean)
```

We will create a panel plot consisting of two plots.  One plot shows average number of steps taken at each 5 minute interval on weekdays and the other does the same for weekends.  First, we will subset our dataset twice creating two new datasets, one with weekday data and one with weekend data.  Then, we will construct the plots using these datasets.  On the plots, the x-axis labels mean the following: 500 = hour 5, minute 0; 1005 = hour 10, minute 5, etc: 

```{r nineteenthchunk, echo=TRUE}
##dataset with only weekday data
aggdata4 <- aggdata3[aggdata3$type=="weekday",]
##dataset with only weekend data
aggdata5 <- aggdata3[aggdata3$type=="weekend",]
```

```{r twentieth chunk, echo=TRUE, fig.height=8, fig.width=8}
##plot
par(mfrow=c(2,1),mar=c(2,2,2,2))
plot.ts(aggdata4$interval,aggdata4$steps,type="l",xlab="5 Minute Interval",ylab="Average Number of Steps",main="Average Number of Steps For Each 5-Minute Interval on Weekdays")
plot.ts(aggdata5$interval,aggdata5$steps,type="l",xlab="5 Minute Interval",ylab="Average Number of Steps",main="Average Number of Steps For Each 5-Minute Interval on Weekends")
```

From this plot, we can see that the individual was most active in the morning leading up to 9am (900 on the x-axis) on weekdays and was more idle during the daytime.  However, on weekends, the individual was active in the morning as well as the afternoon and even into the evening up to about 8pm (2000 on the x-axis).

The knitr packages was installed to process this markdown file into a md and html file.
